installing packages

```{r}
#install.packages("tidyr")
library(tidyr)
library(ggplot2)
library(dplyr)
```

```{r}
#install.packages("janitor")
library(janitor)
```

data scientists spent 50-80 percent of time mired in more mundane labor of cleaning, transforming and wrangling the data

clean_names(data)

resulting names are unique and conist only of the \_ character, numers, and letters. capitalization preferences can be specified using the case parameter

```{r}
starwars

clean_names(starwars,case = "screaming_snake")
clean_names(starwars,case= "small_camel")
newstarwars <- clean_names(starwars,case="screaming_snake")
newstarwars

clean_names(newstarwars)
newstarwars
newstarwars <- rename(newstarwars,'hair*color' = HAIR_COLOR)
clean_names(newstarwars)
newstarwars
```

clean_names is a shortcut for renaming all the columns

```{r}
StarWarsWomen <- select(filter(arrange(starwars,birth_year),sex=="female"),name,species)
StarWarsWomen

starwarswomen <- filter(starwars,sex == "female")
starwarswomen <- arrange(starwarswomen,birth_year)
starwarswomen <- select(starwarswomen,name,species)
```

pipe

function(data,parameters)

becomes

data\|\> function(parameters)

function2(function(data,parameters),params2)

becomes

data\|\>function(parameters)\|\> function2(params2)

slice\_ functions take rows based on your positions

slice_head(data,n=1) takes first row from each group

slice_tail(data,n=1) last row from each group

slice_min(data,x,n=1) takes the row with the smallest value of column x

slice_max(data,x,n=1) takes the row with the largest value of column x

slice_sample(data,n=1) takes out one random row

you can vary n to select more than one row,

or instead of n= you can use prop = 0.1 to select (e.g.) 10% of the rows in each group

```{r}
slice = slice_head(starwars,n=1)
slice

slice <- slice_max(starwars,height,n=10) #10 tallest star wars characters
slice

slice <- slice_max(starwars,height,n=2, by=species,with_ties = F)
slice
```

table4a

```{r}
table4a

tidy_table4a <- pivot_longer(table4a, cols = c('1999','2000'),
             names_to = "year",
             values_to = "cases")

table4b #shows population data
## how would we pivot table4b to the tidy format
tidy_table4b <- pivot_longer(table4b,cols = c('1999','2000'),
             names_to = "year",
             values_to = "population")
```

pivot wider

```{r}
##pivot wider
pivot_wider(table2,
            names_from = type,
            values_from = count)
```

separate

```{r}

table3

separate(table3,rate,into = c("cases","population"),
         sep = "/")
```

unite

```{r}
table5

tidy_table5 <- unite(table5,
      "year",
       c("century","year"),
      sep = "")

```

bind rows

```{r}
new_data <- data.frame(country = "USA", year = "1999",cases = "1042", population = "2000000")

bind_rows(tidy_table5,new_data)
```

# data analysis: hypothesis testing

roadmap:

-   study design

-   hypothesis testing

-   t test, p values and z scores

-   anova test

-   power and error

goal: be able to design a study, test, etc

study design: indipendent and dependent variables

independent = i change

dependent = data

does the number of hours a student studies influence how well they do on an exam

indipendent: hours of studying dependent: grade on exam

does the type of music a plant is exposed to affect its growth

independent: type of music dependent: growth

is exercising related to happier relationships

independent: exercise dependent: happy relationship

correlation does not = causation. if two things go up at the same time dosent mean they're connected

we do things on a sample group and how we sample things affects our study

convenience sample: people who are most nearby to you

random sample: randomly choosing people to survey

cluster sample: groups of people in different clusters

stratified sample: different strata, random sample in each strata (for example different grade levels)

normal distribution: bell curve shape, mean and median are same, 99.7% of data within 3 standard deviations of mean/median

z value = value-mean/standard deviation

if positive means greater than mean, if negative then smaller than mean

using a sample we can try to draw conclusions on whats true of the general population

null hypothesis: hypothesis if theres no change

alternate hypothesis: what your actual hypothesis is

hypothesis testing:

-   determining the null and alternative hypothesis

-   determining the direction of the test

-   calculating the z-value (how many standard deviations away from mean) based on your observation

-   find the p value (probability of this happening by chance)

-   comparing the p value to significant level

one tailed:

alternative hypothesis: great than or less than

two tailed:

alternative hypothesis: not equal to

for example: we think there are less than 40 cookies so we do one tailed

p value: probability of obtaining the observed difference if the null hypothesis is true

how do we calculate p values

-   to calculate a p value, we first have to calculate a test statistic

-   the test statistic is a number calculated from a statistical test of a hypothesis

-   it shows how closely hour observed data match the distribution expected under the null hypothesis of that statistical test

how do we calculate test statistic?:

depends on the test we are using

we are going to start with the one sample t test

we will need:

-   the difference we are testing

-   the spread of the data in out sample

-   the amount of data in our sample

for our sample we will need

the difference we are testing: sample mean = 6.5, population mean=7

the spread of the data in our sample: sample standard deviation = 1.35

amount of data in our sample = 10

t = 6.5-7/1.35/sqr10

to convert to a p value

t.test(sample_data,mu=pop_mean), this gives us a p value

p value = 0.279

there is a 27.29% chance we would see a difference of 0.5 or more just by chance

more than 1 out of 4 times we say theres a difference, there isnt actually a difference

25% of the time is far too often to say that this is most likely a real difference

the widely accepted rare enough percentage is 5%

if p value\<0.05, the data supports the alternate hypothesis, there is a significant difference

if p value \>= 0.05, the data fails to reject the null hypothesis, there is not a significant difference

we want to know is there a relationship between coffee drinking and the number of hours of sleep a student gets each night

null hypothesis: no relationship

Alternate hypothesis: there is a relationship.

t.test(sample1_data, sample2_data)

sample1_data is the data for your first category as a numeric vector (the hours of sleep for al coffee drinkers)

sample 2_data is the data for your second category as a numeric vector (th ehours of sleep for all non coffee drinkers)

other types of t tests:

one sample, one tailed test (t.test(sample_data,mu = mean,alternative = "greater")

two sample, one tailed test (t.test(sample1_data,sample2_data,alternative="less")

two sample, paired, two tailed test:

t.test(sample1_data,sample2_data,paired = TRUE

one sample: is there a difference betwene a group and the population

indepdnent sample: is there a difference betwene two groups

paired samples: is there a difference in a group between two points in time
