---
editor_options: 
  markdown: 
    wrap: 72
---

5 main dplyr functions

1.filter(): subset rows in a dataset based on data values 2.arrange():
reorder the rows 3.select(): subset columns in a dataset based on column
names 4.mutate(): create new variables with functions of existing
variables 5.summarize(): collapse many values down to a single summary

1.  first argument is a dataframe
2.  following arguments describe what to do with the data frame using
    the column names
3.  result is a new dataframe

```{r setup}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)

```

we will be practicing with dplyr functions using the `starwars`dataset

```{r}
head(starwars)
?starwars
```

## subset rows with filter

filter() allows us to subset the rows of a dataset based on the values

```{r}
filter(starwars,species == "Human")

##save to a new R object
humans <- filter(starwars, species == "Human")
humans
filter(humans, sex == "male")

#3filter multiple columns at once
filter(starwars, species == "Human", sex== "male")
```

we can use different comparison operators to filter

-   `!`is not

-   `==`is equal

-   `&` is "and"

-   `|`is or

-   `<`is less than

-   `>`is greater than

-   `is.na()`returns true is value is NA

```{r}
#3return characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
#how to get all characters that arent human
filter(starwars, species!= "Human")
```

variable_name %in% vector_of_values returns TRUE if the value of
variable_name is equal to one of the values in vector_of_values

this is a shortcut for variable_name ==
vector_of_values[1]\|variable_name == vector_of_values[2]

```{r}
"yellow" %in% c("yellow","fair","tan")
skin_color <- "grey"
skin_color %in% c("yellow","fair","tan")

## how would we filter for characters with brown of blond hair?

filter(starwars,hair_color %in% c("blond","brown"))
```

grepl("characters_of_interest",variable_name) returns TRUE if the value
of variable_name contains the characters_of_interest

```{r}
filter(starwars,grepl("blond",hair_color))

#3how would we get all characters with blue in their eyes

filter(starwars, grepl("blue",eye_color))
```

you can use \<, \>, \<=, \>= to compare numeric variables or categorical
variables

-   nominal categorical variables (character vectors), are ranked
    alphabetically

-   ordinal categorical variables (facotrs) are ranked based on the
    built in order

```{r}
##starwars characters with a name starting with letters before #D
filter(starwars, name<"D")

str(diamonds)
filter(diamonds,cut>"Good")

"D" < "d" #false
"d" < "D" #true
```

between(x,left,right) is a shorthand for x\>= left and x\<= right

```{r}
between(5,1,7) #true bc 5 is less than or equal to 7 and greater than or equal to 1
filter(starwars,between(mass,50,100))
```

practice!

find all starwars characters that

1.are missing a hair color

2.  are shorter than 100cm and are not droids
3.  born between the years 100 and 200
4.  another of your choice

```{r}
filter(starwars,is.na(hair_color)) #1
filter(starwars,height<100 & species != "Droid") #2
filter(starwars,between(birth_year,100,200)) #3
filter(starwars,between(birth_year,19,32)& species == "Human")#4
```

# order rows with arrange

arrange() reorders rows based on the values of columns(s)

```{r}
##order by character names
arrange(starwars,name)
# order of character values: aAbBcCdD

#3order by species and birth year
arrange(starwars,species, birth_year)

## use desc() to get the birth years in descending order
arrange(starwars,species,desc(birth_year))
```

practice!

1.arrange the starwars dataset from tallest to shortest hcaracters

2.  arrange the starwars dataset from shortest to tallest by species
3.  alphabetize the star wars characters by their hair color then eye
    color
4.  your choice

```{r}
#1
arrange(starwars,desc(height))
#2
arrange(starwars,height, species)
#3
arrange(starwars,hair_color,eye_color)
#4
arrange(starwars,birth_year,skin_color,eye_color)
```

## subset columns with `select()`

```{r}
select(starwars,hair_color,skin_color,eye_color)
character_colors <- select(starwars,hair_color,skin_color,eye_color)
character_colors
```

we can also select multiple columns in a row

```{r}
#all columns between name and eye_color inclusive
select(starwars,name:eye_color)
```

we can also remove columns by name

```{r}
select(starwars,-films,-vehicles,-starships)
select(starwars,-c(films,vehicles,starships))
select(starwars,-c(films:starships))
```

everything() is a shortcut for "everything else")

```{r}
select(starwars,species,everything())#put species first then everything else

#3alphabatize columns
starwars_names <-names(starwars)
select(starwars,starwars_names)
```

starts_with("abc") matches column names that start with "abc"

ends_with("abc")

contains("abc") matches column names that contain "abc"

```{r}
select(starwars,starts_with("h"))

select(starwars,contains("or"))
```

practice

1.  select each characters homeworld and species
2.  brainstorm as many ways as possible to select
    hair_color,skin_color,eye_color and birth_year

```{r}
select(starwars,homeworld,species)
```

```{r}
#1
select(starwars,hair_color,skin_color,eye_color,birth_year)
#2
select(starwars,hair_color:birth_year)
#3
select(starwars,contains("_"))
#4
select(starwars,contains("color"),birth_year)
#5
select(starwars,-c(name:mass,sex:starships))
```

add new variables with mutate

```{r}
starwars_small <-select(starwars,height,mass,species)
starwars_small

mutate(starwars_small,height_m = height/100)
```

we can also creat multiple variables at once and we can refer to
variables we just created

```{r}
mutate(starwars_small,height_m = height/100,bmi = mass/height_m^2)

#3only keep new variables
transmute(starwars_small,height_m = height/100, bmi = mass/height_m^2)
```

summary statistics give one value for the entire column: sum(),mean()

```{r}
mutate(starwars_small,sum=sum(mass,na.rm=T))

mutate(starwars_small,row_addition = height+mass,summary_sum = sum(height,mass,na.rm = T))
```

group by a variable using the .by parameter

```{r}
mutate(starwars_small,
       species_mass = sum(mass,na.rm = T),
       .by = species)
```

practice

1.convert mass to pounds (1kg = 2.2lbs

```{r}
mutate(starwars_small,
       pounds = mass*2.2)
```

change the type of a variable

```{r}
mutate(starwars_small,
       mass = as.logical(mass))
```

### summarize data with summarize

```{r}
summarize(starwars,mean_mass = mean(mass,na.rm=T))

#group by species
summarize(starwars,mean_mass = mean(mass,na.rm=T),
          .by = species)
```

the only columns kept are the columns you group by and the columns you
create

n() is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars,mean_mass = mean(mass,na.rm = T),
          count = n(),
          .by =species)

#3group by multiple groups
summarize(starwars,mean_mass = mean(mass,na.rm = T),
          count = n(),
          .by =c(species,sex))
```

### renaming

rename(data,new_name = old_name)

```{r}
rename(starwars,
       birthYear = birth_year)
#dosent change the original dataset
```
